Gringo 2.0

### 2008-07-30 doc/documentation/paper.tex
changes according to sven, unfinished
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@413 3e02f596-3131-0410-8352-e04eeaec37cc

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. First, the general restrictions to the input language are stated. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured, less restricted input language.
gringo2.0 supports several output formats and even a human readable format for debugging purposes.
It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

\section{$\lambda$-Restriction}
\label{domainrestricted}
Gringo verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. The definition of lambda restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.

\section{Interface}
Gringo can be started on the command line and has some parameters that can be useful.
The first parameters are gringo specific parameters and any arbritary number of filenames of the logic program.
\paragraph{--help}
Gives a short overview of all parameters.
\paragraph{--const}
You can set the value of constants defined in your logic program via the command line of gringo.
After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant. The integer value can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the atom $n$ is replaced by $10$ in the logic program.

\paragraph{--convert}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{-l}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{-c}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore adouble minus ``--'' must be given after all gringo related parameters are set. After them die clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{-p}
Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
\paragraph{-g [1-7]}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The optional number can range from $1$ to $7$ and refers to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normalform is specified, conglomeration(7)  is assumed.

\paragraph{default parameters}
The default parameter is ``-l'' for smodels output.
\section{Language Constructs}
The basic elements of a logic program are atoms, constants, variables and rules.
\paragraph{Constant}
A constant is a representation of an entity in the problem domain.
A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
It also can be encompassed by double quotation marks. When using quotation marks, which is sometimes helpful when using constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
\paragraph{Variable}
A variable is programming construct to simplify writing logic programs.
It always starts with a capitalized letter and can contain digits and letters.
In the grounding process a variable is replaced by one or more constants.
Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
\paragraph{Atom}

An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
It has to start with a lower case letter or an underscore.
An atom can have several parameters, encapsulated by parenthesis and separated with commas.
The parameters are either constants, variables or function symbols.
Function symbols do have the same syntax as atoms. This way they can be finitely nested.
As an example, an atom with three parameters is given:
The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}

\paragraph{Rule}
A rule normally consists of a head and a body and ends with a dot.
The body is separated from the head using ``:-'' as in
\begin{verbatim}
 head :- body.
\end{verbatim}
The body is a comma separated conjunction of expressions.
The supported expressions are introduced in the next section.
The variables used in the head must be restricted to a domain in the body.
Check (\ref{domainrestricted}) to see the restrictions on domains.
\newline
A rule can occur without a body as a fact, therefore the body is omitted.
\begin{verbatim}
 head.
\end{verbatim}
The same can be done with the body to express an integrity constraint.
\begin{verbatim}
 :- body.
\end{verbatim}

\paragraph{Comments}
Comments included in the input are line based.
Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.
\paragraph{Warnings}
Several warnings are produced by gringo to give hints on common mistakes.
They are all printed to standard error output.

\paragraph{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of predicates that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format.
\begin{verbatim}
 "Disjunction not allowed on this normalform."
 "Count aggregate/cardinality constraint not allowed
  in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Max aggregate not allowed in this normalform."
 "Max aggregate not allowed
  with negative weights in this normalform."
 "Min aggregate not allowed in this normalform."
 "Min aggregate not allowed 
  with negative weights in this normalform."
 "Optimize statement not allowed in this normalform."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normalform."
% "Times aggregate not allowed
%  with negative weights in this normalform."
%\end{verbatim}
These errors belong to the ASPils output format. Another normal form as to be chosen.
\end{itemize}

\subsection{Meta Constructs}
The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
\paragraph{hide}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``name''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify with other variables. For example hide p(X,X) would also hide the fact p(a,b). The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
\begin{verbatim}
 #hide.
\end{verbatim}
in conjunction with the ``show'' statement is used.
The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}
The example hides all atoms except the predicate $p$ with arity $3$.
\paragraph{const}
Within the ``const'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``T'' is always bound to the domain of ``time''.
This has the same effect as adding the domain ``time(T)'' to every rule the variable ``T'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
Is has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
The compute statement is solver dependent and supported by every solver.

@@

If we give the problem description and instance to the grounder, it will generate a variable free version of the problem. This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
\newline
To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
\subsection{Features}
The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Expanding Terms}
\subparagraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

@@

\paragraph{Arithmetic Expressions}
The build-in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported. The bitwise operators are executed on signed integers using the binary complement for negative numbers.
Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

@@

@@

\input{benchmark2.result}
\appendixname
\begin{appendix}
\section{EBNF Syntax}
\label{appendix}
\input{appendix}
\end{appendix}

>>>>

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. First, the general restrictions to the input language are stated. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
gringo2.0 supports several output formats and even a human readable format for debugging purposes.
It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

\section{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
Gringo verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.

\section{Interface}
Gringo can be started on the command line and has some parameters that can be useful.
The first parameters are gringo specific parameters and consist of a number of options and any arbritary number of filenames of the logic program. For the special purpose that Gringo is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the Gringo parameters.
\paragraph{\texttt{--help}}
Gives a short overview of all parameters.
\paragraph{\texttt{--const}}
You can set the value of constants defined in your logic program via the command line of gringo.
After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant. The integer value can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the atom $n$ is replaced by $10$ in the logic program.

\paragraph{\texttt{--convert}}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{\texttt{-l}}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{\texttt{-c}}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore adouble minus ``\texttt{--}'' must be given after all gringo related parameters are set. After them die clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{\texttt{-p}}
Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
\paragraph{\texttt{-g [1-7]}}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The optional number can range from $1$ to $7$ and refers to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normalform is specified, conglomeration(7)  is assumed.

\paragraph{default parameters}
The default parameter is ``\texttt{-l}'' for smodels output.
\section{Syntax}
The basic elements of a logic program and its syntax is now described.
Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
\begin{itemize}
\item a set P of predicate symbols,
\item a set F of function symbols,
\item a set V of variables,
\item a set C of constants,
\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
\end{itemize}

TODO: Which greek letters for P, F, L and C
\paragraph{Predicates and function symbols}
Predicates and function symbols have a predicate symbol/name and an arity.
For the function symbol $f$ with arity 3 we write $f/3$.
The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
With $P$ we denote to the set of predicates and $F$ denotes the set of function symbols.
\paragraph{Constants}
Constants are function symbols with an arity of zero.
The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces. Tt has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not equal to $a$.
$C$ is the set of constants
\paragraph{Variables}
Variables have to start with a capitalized letter and can contain digits and letters.
The set of variables is denoted by $V$.
The language over $P$, $F$and $C$ is denoted by $L(P,F,C)$.
\paragraph{Terms}
The building blocks of the language are terms.
Each constant $c \in C$ and each variable $f \in F$ is a term.
Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in F$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
Since a term is defined inductively, it can be nested.
A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}
\paragraph{Atoms}
An atom over $L(P, F, C)$ is of the form $p(t1 , . . . , tn )$ where $p/n$ is a predicate
symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
\paragraph{Rules}
A rule over the language $L(P,F,C)$ has the form
$$
A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
$$
where $A_0$ to $A_n$ are atoms. \newline
A normal logic program $\Pi$ over $L(P,F,C)$ is a finite set of rules over $L(P,F,C)$.

%\paragraph{Function Symbols}
%A constant is a representation of an entity in the problem domain.
%A constant is either a name, starting with a lower case letter or an underscore, or a positive integer.
%It also can be encompassed by double quotation marks. When using quotation marks, which is sometimes helpful when using %constants with spaces, it has to be remarked that a constant with quotation marks is never equal to a constant without, so ``$a$'' is not $a$.
%\paragraph{Variable}
%A variable is programming construct to simplify writing logic programs.
%It always starts with a capitalized letter and can contain digits and letters.
%In the grounding process a variable is replaced by one or more constants.
%Each variable occurring in a rule must be domain restricted as described in section \ref{domainrestricted}.
%\paragraph{Atom}
%An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
%It has to start with a lower case letter or an underscore.
%An atom can have several parameters, encapsulated by parenthesis and separated with commas.
%The parameters are either constants, variables or function symbols.
%Function symbols do have the same syntax as atoms. This way they can be finitely nested.
%As an example, an atom with three parameters is given:
%The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
%\begin{verbatim}
% p(a,f(b,X), g(f(X),Y)).
%\end{verbatim}

%\paragraph{Rule}
%A rule normally consists of a head and a body and ends with a dot.
%The body is separated from the head using ``:-'' as in
%\begin{verbatim}
% head :- body.
%\end{verbatim}
%The body is a comma separated conjunction of expressions.
%The supported expressions are introduced in the next section.
%The variables used in the head must be restricted to a domain in the body.
%Check (\ref{domainrestricted}) to see the restrictions on domains.
%\newline
%A rule can occur without a body as a fact, therefore the body is omitted.
%\begin{verbatim}
% head.
%\end{verbatim}
%The same can be done with the body to express an integrity constraint.
%\begin{verbatim}
% :- body.
%\end{verbatim}

\subsection{Meta-Syntax}
The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
\paragraph{hide}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``name''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify with other variables. For example hide p(X,X) would also hide the fact p(a,b). The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
\begin{verbatim}
 #hide.
\end{verbatim}
in conjunction with the ``show'' statement is used.
The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}
The example hides all atoms except the predicate $p$ with arity $3$.
\paragraph{const}
Within the ``const'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``T'' is always bound to the domain of ``time''.
This has the same effect as adding the domain ``time(T)'' to every rule the variable ``T'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
Is has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
The compute statement is solver dependent and supported by every solver.
\paragraph{Comments}
Comments included in the input are line based.
Every line starting with ``\%'' is a comment. The whole line is omitted and not to be handled by the grounder.

@@

If we give the problem description and instance to the grounder, it will generate a variable free version of the problem. This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
\newline
To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
\subsection{Features}
The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Function Symbols}
Function symbols are fully supported with Gringo2.0.
They can be finitly nested and also be used as domain predicates.
While in lParse function symbols are allowed in a very restricted way, no restrictions apply in Gringo2.0.
The following program can not be grounded with lParse since $X$ is an unrestricted variable. With Gringo2.0 the program is now groundable.
\begin{verbatim}
a(X) :- f(1,g(X)).
f(1,X) :- f(X).
f(g(2)).
\end{verbatim}
From the second rule the domain \texttt{f(1,g(2))} is generated.
This domain can be used for the first rule and results in the grounded rule
\begin{verbatim}
 a(2) :- f(1,g(2)).
\end{verbatim}

\paragraph{Expanding Terms}
\subparagraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

@@

\paragraph{Arithmetic Expressions}
The built-in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported. The bitwise operators are executed on signed integers using the binary complement for negative numbers.
Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

@@

\subsection{Warnings}
Several warnings are produced by gringo to give hints on common mistakes.
They are all printed to standard error output.

\subsection{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of predicates that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format.
\begin{verbatim}
 "Disjunction not allowed on this normalform."
 "Count aggregate/cardinality constraint not allowed
  in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Max aggregate not allowed in this normalform."
 "Max aggregate not allowed
  with negative weights in this normalform."
 "Min aggregate not allowed in this normalform."
 "Min aggregate not allowed 
  with negative weights in this normalform."
 "Optimize statement not allowed in this normalform."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normalform."
% "Times aggregate not allowed
%  with negative weights in this normalform."
%\end{verbatim}
These errors belong to the ASPils output format. Another normal form as to be chosen.
\end{itemize}

@@

\input{benchmark2.result}
\section{Future Work}
$\Pi$-restrictedness
bottom up grounding
\newpage
\textbf{\appendixname}
\begin{appendix}
\section{EBNF Syntax}
\label{appendix}
\input{appendix}
\end{appendix}
